Supplementary materials: $R^*$ convergence diagnostic

### 2020-09-29 src/s_discrete.Rmd
added discrete example

P <- matrix(nrow = 4, ncol = 4)
P[1, ] <- c(0, 1/2, 1/2, 0)
P[2, ] <- c(1/2, 0, 1/3, 1/6)
P[3, ] <- c(1/4, 1/4, 1/4, 1/4)
P[4, ] <- c(1/2, 0, 0, 1/2)

@@

P0 <- P
P1 <- matrix(nrow = 4, ncol = 4)
P1[1, ] <- c(0, 1/2, 1/2, 0)
P1[2, ] <- c(1/2, 0, 1/3, 1/6)
P1[3, ] <- c(1/4, 1/4, 1/4, 1/4)
P1[4, ] <- c(1/2, 0, 1/6, 1/3)

@@

nreplicates <- 20
Ls <- c(1000, 2000, 5000, 10000)

# create P1 matrices
f_create_P1_example <- function(last_val=1/2){
  P1 <- matrix(nrow = 4, ncol = 4)
  P1[1, ] <- c(0, 1/2, 1/2, 0)
  P1[2, ] <- c(1/2, 0, 1/3, 1/6)
  P1[3, ] <- c(1/4, 1/4, 1/4, 1/4)
  P1[4, ] <- c(1/2, 0, (1 - 1/2 - last_val), last_val)
  return(P1)
}

last_vals <- c(0, 0.25, 0.5)
P1s <- map(last_vals, ~f_create_P1_example(.))

@@

Plot
```{r}
a_sum <- r_star_vals %>%
  group_by(P1, `sample size`, type) %>%
  summarise(lower=quantile(value, 0.025),
         upper=quantile(value, 0.975),
         middle=quantile(value, 0.5))
r_star_vals %>% 
  left_join(a_sum) %>% 
  ggplot(aes(x=as.factor(`sample size`), y=value)) +
  geom_jitter(width = 0.2, colour="grey") +
  geom_pointrange(aes(ymin=lower, y=middle, ymax=upper)) +
  facet_grid(vars(P1)) +
  geom_hline(yintercept = 1, linetype=2) +
  scale_y_continuous(limits=c(0.9, 1.2))
```

>>>>

P <- matrix(nrow = 4, ncol = 4)
P[1, ] <- c(0, 1/2, 1/2, 0)
P[2, ] <- c(1/2, 0, 1/3, 1/6)
P[3, ] <- c(1/4, 1/4, 1/4, 1/4)
P[4, ] <- c(0, 1, 0, 0)

@@

P0 <- P
P1 <- matrix(nrow = 4, ncol = 4)
P1[1, ] <- c(0, 1/2, 1/2, 0)
P1[2, ] <- c(1/2, 0, 1/3, 1/6)
P1[3, ] <- c(1, 0, 0, 0)
P1[4, ] <- c(1, 0, 0, 0)

@@

nreplicates <- 40
Ls <- c(1000, 2000, 5000, 10000)

# create P1 matrices
f_create_P1_example <- function(per_along){
  P1 <- matrix(nrow = 4, ncol = 4)
  P1[1, ] <- c(0, 1/2, 1/2, 0)
  P1[2, ] <- c(1/2, 0, 1/3, 1/6)
  v_diff <- c(1, 0, 0, 0) - c(1/4, 1/4, 1/4, 1/4)
  P1[3, ] <- c(1/4, 1/4, 1/4, 1/4) + v_diff * per_along
  v_diff <- c(1, 0, 0, 0) - c(0, 1, 0, 0)
  P1[4, ] <- c(0, 1, 0, 0) + v_diff * per_along
  return(P1)
}

last_vals <- c(0, 0.5, 1)
P1s <- map(last_vals, ~f_create_P1_example(.))

@@

Plot
```{r}
temp <- readRDS("../data/discrete_r_star_hat.rds")
r_star_vals <- temp$r_star %>% 
  mutate(P1=if_else(P1==1, "P1", if_else(P1==2, "P2", "P3"))) %>% 
  mutate(`sample size`=as.character(`sample size`)) %>% 
  mutate(`sample size`=if_else(`sample size`=="10000", "10,000", `sample size`)) %>% 
  mutate(`sample size`=as.factor(`sample size`)) %>% 
  mutate(`sample size`=fct_relevel(`sample size`, "1000", "2000", "5000", "10,000"))
a_sum <- r_star_vals %>%
  group_by(P1, `sample size`, type) %>%
  summarise(lower=quantile(value, 0.025),
         upper=quantile(value, 0.975),
         middle=quantile(value, 0.5))
g1 <- r_star_vals %>% 
  left_join(a_sum) %>% 
  ggplot(aes(x=as.factor(`sample size`), y=value)) +
  geom_jitter(width = 0.2, colour="grey") +
  geom_pointrange(aes(ymin=lower, y=middle, ymax=upper)) +
  facet_grid(vars(P1)) +
  geom_hline(yintercept = 1, linetype=2) +
  scale_y_continuous(limits=c(NA, NA)) +
  xlab("Sample size") +
  ylab(TeX("$R*$")) +
  theme(axis.text = element_text(colour="black", size=16),
        axis.title = element_text(colour="black", size=16),
        strip.text.y = element_text(colour="black", size=16),
        legend.position = "none",
        title = element_text(colour="black", size=16)) +
  ggtitle("A.")
r_star_vals <- temp$r_hat %>% 
  mutate(P1=if_else(P1==1, "P1", if_else(P1==2, "P2", "P3"))) %>% 
  mutate(`sample size`=as.character(`sample size`)) %>% 
  mutate(`sample size`=if_else(`sample size`=="10000", "10,000", `sample size`)) %>% 
  mutate(`sample size`=as.factor(`sample size`)) %>% 
  mutate(`sample size`=fct_relevel(`sample size`, "1000", "2000", "5000", "10,000"))
a_sum <- r_star_vals %>%
  group_by(P1, `sample size`, type) %>%
  summarise(lower=quantile(value, 0.025),
         upper=quantile(value, 0.975),
         middle=quantile(value, 0.5))
g2 <- r_star_vals %>% 
  left_join(a_sum) %>% 
  ggplot(aes(x=as.factor(`sample size`), y=value)) +
  geom_jitter(width = 0.2, colour="grey") +
  geom_pointrange(aes(ymin=lower, y=middle, ymax=upper)) +
  facet_grid(vars(P1)) +
  geom_hline(yintercept = 1, linetype=2) +
  scale_y_continuous(limits=c(NA, NA)) +
  xlab("Sample size") +
  ylab(TeX("Split-$\\hat{R}$")) +
  theme(axis.text = element_text(colour="black", size=16),
        axis.title = element_text(colour="black", size=16),
        strip.text.y = element_text(colour="black", size=16),
        legend.position = "none",
        title = element_text(colour="black", size=16)) +
  ggtitle("B.")

multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}

pdf("../output/discrete.pdf", width = 12, height = 8)
multiplot(g1, g2, cols = 1)
dev.off()
```

Plotting stationary distributions
```{r}
m_stationary <- matrix(c(11/46, 15/46, 14/46, 6/46, 71/198, 17/66, 10/33, 8/99, 4/9, 2/9, 8/27, 1/27), nrow = 3, byrow = T) %>% 
  as.data.frame() %>% 
  mutate(matrix=c("P1", "P2", "P3"))
colnames(m_stationary)[1:4] <- c("1", "2", "3", "4")
g <- m_stationary %>% 
  melt(id.vars="matrix") %>% 
  ggplot(aes(x=as.factor(variable), y=value, fill=as.factor(matrix))) +
  geom_col(position = position_dodge()) +
  xlab("State") +
  ylab("Probability") +
  scale_fill_brewer("Transition\nprobability\nmatrix", palette = "Dark2", labels=unname(TeX(c("P_1", "P_2", "P_3")))) +
  theme(axis.text = element_text(colour="black", size=16),
        axis.title = element_text(colour="black", size=16),
        strip.text.x = element_text(colour="black", size=16),
        legend.text = element_text(colour="black", size=16),
        legend.title = element_text(colour="black", size=16),
        legend.position = c(0.85, 0.7))
ggsave("../output/discrete_stationary.pdf", g, width = 10, height = 6)
```

### 2020-09-29 write_up/arxiv_submission_aug_2020/supplementary_materials.tex
added discrete example

>>>>

\section{Discrete distribution}\label{sec:discrete}
In this section, we compare the performance of $R^*$ and $\widehat{R}$ on a target distribution with a discrete dimension. To do so, we generate draws from four discrete Markov chains, following a 1st order Markov process with transition probabilities,
%
\begin{equation}
p_{ij} = \text{Pr}(X_{n+1}=j|X_n=i),
\end{equation}
%
where $p_{ij}$ is the probability of transitioning from state $i\in\{1,2,3,4\}$ at time $n$ to state $j\in\{1,2,3,4\}$ at time $n+1$, and $X(0)=1$. The target distribution is, hence, the stationary distribution of this process.

In three of the four chains, we generate draws using a transition probability matrix:
%
\begin{gather}
\boldsymbol{P}
=
\begin{bmatrix}
0 & 1/2 & 1/2 & 0\\
1/2 & 0 & 1/3 & 1/6\\
1/4 & 1/4 & 1/4 & 1/4\\
0 & 1 & 0 & 0
\end{bmatrix},
\end{gather}
%
which implies a stationary target distribution with the following probability simplex $\pi_1 = (11/46, 15/46, 14/46, 6/46)$.

For the fourth chain, we specify differing transition probability matrices in each of three sets of results. In the first, we assume $\boldsymbol{P}_1=\boldsymbol{P}$ (i.e. the same as used for the other three chains); in the second and third cases, we assume this transition matrix is given by:
%
\begin{gather}
\boldsymbol{P}_2
=
\begin{bmatrix}
0 & 1/2 & 1/2 & 0\\
1/2 & 0 & 1/3 & 1/6\\
5/8 & 1/8 & 1/8 & 1/8\\
1/2 & 1/2 & 0 & 0
\end{bmatrix}\quad
\boldsymbol{P}_3
=
\begin{bmatrix}
0 & 1/2 & 1/2 & 0\\
1/2 & 0 & 1/3 & 1/6\\
1 & 0 & 0 & 0\\
1 & 0 & 0 & 0
\end{bmatrix}.
\end{gather}
%
These correspond to stationary probability simplices $\pi_2 = (71/198, 17/66, 10/33, 8/99)$ and $\pi_3 = (4/9, 2/9, 8/27, 1/27)$. The three stationary distributions corresponding to each transition probability matrices are shown in Fig. \ref{fig:discrete_stationary}. This illustrates that there is an ordering among the stationary distributions $\pi_1\sim\pi_2\sim\pi_3$, meaning $\pi_1$ is most similar to $\pi_2$, and $\pi_3$ is most similar to $\pi_2$. Put another way, $\pi_3$ is most different from $p_1$.

\begin{figure}[!htb]
	\centerline{\includegraphics[width=1.0\textwidth]{discrete_stationary.pdf}}
	\caption{\textbf{Discrete distribution example: stationary distributions.} The three colours illustrate the stationary distributions of the Markov chains corresponding to the matrices $P_1$, $P_2$ and $P_3$ defined in \S\ref{sec:discrete}.}
	\label{fig:discrete_stationary}
\end{figure}

For each of these three cases, we consider a range of sample sizes across each of the four chains: 1000, 2000, 5000 and 10,000. At each combination of transition probability matrix (for the fourth chain) and sample size, we perform 40 replicates. In each replicate, we calculate both $R^*$ and rank-normalised split-$\widehat{R}$. The results of these experiments are shown in Fig. \ref{fig:discrete}. In the horizontal axes of both panels, we show the sample size. The vertical axis indicates the value of $R^*$ and split-$\widehat{R}$ in panels A and B, respectively. The stacked sub-panels within A and B indicate the transition matrix used for the fourth chain. Since $\boldsymbol{P}_1=\boldsymbol{P}$, the stationary distribution for the other three chains, both diagnostics should tend towards indicating convergence in these cases, and, for both statistics, this was typically the case. For $R^*$, the replicates were centred on $R^*=1$, although the stochasticity of replicates meant that, in some cases, $R^*>1$. Whereas, $\widehat{R}<1$ in all cases. Considering the simulations using $\boldsymbol{P}_2$ for the fourth chain transition probability matrix, $R^*>1$ across a majority of replicates, and this majority increased with sample size. In this case, $\widehat{R}<1$ for all replicates across the various sample sizes. For simulations using $\boldsymbol{P}_3$ for the fourth chain transition probability matrix, $R^*>1$ across all replicates and the magnitude of median $R^*$ increased with sample size; as for the $\boldsymbol{P}_2$ case, $\widehat{R}<1$, indicating convergence.

This example hints that $R^*$ may be particularly adept at detecting poor convergence in small discrete parameter spaces.

\begin{figure}[!htb]
	\centerline{\includegraphics[width=1.0\textwidth]{discrete.pdf}}
	\caption{\textbf{Discrete distribution example: convergence diagnostics.} On the horizontal axis, we show the sample size of each set of simulations. The panels within A. and B. indicate the transition probability matrix used for the fourth chain (the other three chains always used $\boldsymbol{P}=\boldsymbol{P}_1$). In panel A, grey points indicate $R^*$ from each replicate calculated using Algorithm \ref{alg:R_star}) , and the black point-ranges indicate 2.5\% , 50\% and 97.5\% quantiles across the 40 replicates. In panel B, we show the same, but for split-$\widehat{R}$. The dashed horizontal lines at 1 indicate the convergence threshold in both cases.}
	\label{fig:discrete}
\end{figure}
